MATH-469 Project Checkin

### 2012-11-22 checkin/project-checkin.tex
checkin overdue commits

The goal of my project is to implement an algorithm that can assemble a genome,
given a set of reads that came from it.  The key idea in the algorithm I am
implementing is the construction of a string graph where the edges are labeled
with DNA sequences and the vertices correspond to reads, and a path through the
graph corresponds to a set of reads that assemble together consistently.

I have implemented the initial stages of the algorithm.  Each step has been
implemented as a separate binary program.  This means that the overall assembly
process will iteratively transform the data (including the graph) on-disk into
the final assembly.  The main currently implemented programs are:

@@

\item
	{\tt remove-contained-reads}:
	Given a set of reads and the overlaps that were computed from them,
	finds all reads that are fully contained by another read and discard
	them, along with the corresponding overlaps.  If a pair of reads is
	identical, only one copy of the read is kept.

@@

@@

The modular design allows for multiple possible data flows, but a possible flow
is
%{\tt convert-reads} $\to$
{\tt compute-overlaps} $\to$ {\tt
remove-contained-reads} $\to$ {\tt build-directed-string-graph} $\to$ {\tt
transitive-reduction} $\to$ {\tt collapse-unbranched-paths} $\to$ {\tt digraph-to-bidigraph}, which will produce a
transitively-reduced, collapsed bidirected string graph produced from overlaps
between non-contained reads.

Even with the final stages not implemented, the algorithm is already very
effective when used on completely random genomes because, since there are no
spurious overlaps in a completely random genome with adequate length reads, the
transitively-reduced string graph is a chain of vertices with no branching,
which then collapses into a single edge that will be labeled with almost all the
original sequence.
%This is true even if the genome is 15 million random base pairs, given 100 bp
%reads sampled uniformly from either strand with no errors.  (Such a genome
%takes about 1 minute to assemble with maximum memory usage 1.8 GB, with the
%main time and memory bottleneck being the {\tt compute-overlaps} program).
However, genomes with repeats will produce tangled graphs that will require the
network flow and Eulerian path parts of the overall algorithm to be implemented.

What follows is an explanation of the initial stages of the algorithm on a very
simple case.  The diagram below shows a 64-base-pair ``genome'' that has had
three 32-base-pair error-free reads taken from it.  Note that the genome is
two-stranded.  {\tt A} always pairs with {\tt T} and {\tt C} always pairs with
{\tt G}, but the two strands run in opposite directions.  Reads 1 and 2 come
from the top (reverse strand), while read 3 comes from the bottom (forward
strand).  The reads are directed as shown (e.g.  read 1 begins with {\tt ACAGC},
reading from right to left, not {\tt TGAGC}).

@@

From looking at the above diagram, it is apparent that read 1 overlaps with read
2 by 20 base pairs, read 2 overlaps with read 3 by 20 base pairs, and read 1
overlaps with read 3 by 8 base pairs.  The task of the {\tt compute-overlaps}
program is to determine these overlaps.  Note that an overlap may match either
the forward or reverse-complement sequence.

@@

Given the set of overlaps, the next step is to build the bidirected or directed
string graph.  Both these graphs are shown on the next page.  In the directed
graphs, a node $n.B$ means the beginning of read $n$, while a node $n.E$ means
the end of read $n$.  In the bidirected graphs, node $n$ simply corresponds to
read $n$.  In both the directed and bidirected graphs, the edges are labeled by
DNA sequences; however, in the drawings, only the lengths of the sequences are
shown.

@@

\begin{figure}
\caption{Transitively-reduced, collapsed bidirected string graph for error-free
100bp reads uniformly sampled from the first 250000 base pairs of the genome of
{\it E. coli}, which includes some repetitive sequences.  The network flow
algorithm must be able to determine how many times each edge should be
traversed, then an Eulerian path through the bidirected graph must be found to
produce a possible assembly.  However, in the general case, the final
reconstruction may have to take the form of a set of paths rather than just one
path.  }
\includegraphics[scale=0.6]{E_coli-crop.pdf}
\end{figure}

>>>>

The goal of my project is to implement an algorithm that can assemble a genome,
given a set of reads that came from it.  The key idea in the algorithm I am
implementing is the construction of a string graph where the vertices correspond
to reads and the edges are labeled with DNA sequences, and a path through the
graph corresponds to a set of reads that assemble together consistently.

I have implemented the initial stages of the algorithm.  Each step has been
implemented as a separate executable program.  This means that the overall
assembly process will iteratively transform the data (e.g. the string graph)
on-disk into the final assembly.  The main currently implemented programs are:

@@

\item
	{\tt remove-contained-reads}:
	Given a set of reads and the overlaps that were computed from them,
	finds all reads that are fully contained by another read and discards
	them, along with the corresponding overlaps.  If a pair of reads is
	identical, only one copy of the read is kept.

@@

\item
	{\tt map-contained-reads}:
	Map the reads that were removed in {\tt remove-contained-reads} back
	into the string graph to mark edges that are supported by more than one
	read and therefore are more likely to need to be traversed multiple
	times.

@@

The modular design allows for multiple possible data flows, but a possible flow
is
%{\tt convert-reads} $\to$
{\tt compute-overlaps} $\to$ {\tt
remove-contained-reads} $\to$ {\tt build-directed-string-graph} $\to$ {\tt
transitive-reduction} $\to$
{\tt map-contained-reads} $\to$
{\tt collapse-unbranched-paths} $\to$ {\tt digraph-to-bidigraph}, which will produce a
transitively-reduced, collapsed bidirected string graph produced from overlaps
between non-contained reads.

Even with the final stages not implemented, the algorithm is already very
effective when used on completely random genomes.  Since there are no spurious
overlaps in a completely random genome with adequate length reads, the
transitively-reduced string graph is a chain of vertices with no branching,
which then collapses into a single edge that will be labeled with almost all the
original sequence.
%This is true even if the genome is 15 million random base pairs, given 100 bp
%reads sampled uniformly from either strand with no errors.  (Such a genome
%takes about 1 minute to assemble with maximum memory usage 1.8 GB, with the
%main time and memory bottleneck being the {\tt compute-overlaps} program).

However, genomes with repeats will produce tangled graphs that will require the
network flow and Eulerian path parts of the overall algorithm to be implemented.

What follows is an explanation of the initial stages of the algorithm on a very
simple case.  The diagram below shows a 64-base-pair ``genome'' (G.) that has
had three 32-base-pair error-free reads taken from it (1., 2., and 3.).  Note
that the genome is two-stranded.  {\tt A} always pairs with {\tt T} and {\tt C}
always pairs with {\tt G}, but the two strands run in opposite directions.
Reads 1 and 2 come from the top (``reverse'') strand, while read 3 comes from
the bottom (``forward'') strand.  The reads are directed as shown (e.g.  read 1
begins with {\tt ACAGC}, reading from right to left.)

@@

From looking at the above diagram, it is apparent that read 1 overlaps with read
2 by 20 base pairs, read 2 overlaps with read 3 by 20 base pairs, and read 1
overlaps with read 3 by 8 base pairs.  The task of the {\tt compute-overlaps}
program is to determine these overlaps.  Note that reverse-complement overlaps
(e.g. reads 2 and 3) are permissible.

@@

Given the set of overlaps, the next step is to build the bidirected or directed
string graph.  For this example, both these graphs are shown on the next page.
In the directed graphs, a node $n.B$ means the beginning of read $n$, while a
node $n.E$ means the end of read $n$.  In the bidirected graphs, node $n$ simply
corresponds to read $n$.  In both the directed and bidirected graphs, the edges
are labeled by DNA sequences; however, in the drawings, only the lengths of the
sequences are shown.

@@

\begin{figure}
\caption{Transitively-reduced, collapsed bidirected string graph for error-free
100bp reads uniformly sampled from the first 250000 base pairs of the genome of
{\it E. coli}, which includes some repetitive sequences.  The network flow
algorithm must be able to determine how many times each edge should be
traversed, then a generalized Eulerian path through the bidirected graph must be
found to produce a possible assembly.  In the general case, the final
reconstruction may have to take the form of a set of paths rather than just one
path.  }
\begin{center}
\includegraphics[scale=0.6]{E_coli-crop.pdf}
\end{center}
\end{figure}
